---
title: "Propensity Score Analysis in R - Advanced"
author: "Statistical Horizons"
date: "October 6-8, 2022"
output:
  pdf_document:
    number_sections: yes
    toc: yes
    toc_depth: 4
header-includes:
- \usepackage{float}
- \usepackage{pdflscape}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage{pgfornament}
- \usepackage{mathtools}
- \usepackage{amsmath,amsthm,amssymb}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \pagenumbering{gobble}
- \usepackage{upgreek}
---

```{r include=F}
options(scipen = 999)
```

\newpage
# Section 3.8: Computer Lab: Running PSS in conjunction with Cox Regression with R

The advantages of propensity score subclasification are: (1) allows analysis of an outcome variable that is virtually any type; and (2) retains a large size of the original sample. The disadvantage is that it is sensitive to the violation of the overlapping assumption.

This lab illustrates propensity score subclassification followed by a multivariate outcome analyis using a Cox proportional hazards model.^[Perkins, S. M., Tu, W., Underhill, M. G., Zhou, X. H., & Murray, M. D. (2000). The use of propensity scores in pharmacoepidemiologic research. *Pharmacoepidemiology and Drug Safety*, 9(2), 93–101. https://doi.org/10.1002/(SICI)1099-1557(200003/04)9:2<93::AID-PDS474>3.0.CO;2-I]

We will demonstrate five steps of subclassification:

1. Sort the sample by estimated propensity scores in an ascending order
2. Divide the sample into $K$ strata using quantiles (quintiles, deciles, etc.)
3. Evaluate the treatment effect by calculating mean difference of outcome and variance of difference between treated and control participants within each stratum, or by running a multivariate analysis of outcomes within each stratum as one does for samples generated by a randomized experiment
4. Estimate the mean difference (ATE) for the whole sample (i.e., all $K$ strata combined) through aggregating
5. Test whether the sample difference on outcome is statistically significant

Finally, we will show an example of marginal mean weighting through stratification (MMWS), which is an alternative method of estimating marginal effects after stratum weighting (as opposed to aggregating the results).

## Load Packages

The `haven` and `sjlabelled` packages are used to load and clean Stata data files (.dta); the `MatchIt` package contains a function for subclassification; and the `cobalt` package contains functions for checking balance.

```{r message=F, warning=F, error=F}
library(haven)
library(sjlabelled)
library(MatchIt)
library(cobalt)
library(tidyverse)
library(broom)
library(survival)
library(PSweight)
library(sandwich)
```

## Import Data

THis dataset contains 2,758 children from the panel data of the National Survey of Child and Adolescent Well-Being (NSCAW). After a listwise deletion of missing data, the study sample comprises 2,723 children. 

The research question is whether caregivers' use of substance abuse services (`aodserv`) is predictive of the timing (`durm`) of a maltreatment re-report (`rrpt`) 18 months after the baseline interview. Of the 2,723 children included in the study, 294 (10.8%) had female caregivers who received substance abuse treatment, and the remaining 2,429 (89.2%) had female caregivers who did not receive treatment services.

```{r}
df <- read_dta("data/adv/sur_subclass.dta") %>%
  haven::zap_formats() %>%
  sjlabelled::remove_all_labels() %>%
  as_tibble()
head(df)
nrow(df)
table(df$aodserv)
```

## Create Propensity Scores Using Logistic Regression

Here we calculate the probability of treatment assignment conditional on the known covariates using logistic regression. We obtain predicted probabilities from the logistic regression model as the propensity scores and add them to our orginal dataframe.

```{r}
(f <- cobalt::f.build("aodserv", select(df, married:cgneed, PSH17A:CRA47A)))
m1 <- glm(f, family = binomial, data = df)
ps <- predict(m1, type = "response")
df.ps <- df %>%
  mutate(ps = ps)
```

## Check Balance Before Subclassification

### Function to Check Balance Using Normalized Difference

```{r}
nd <- function(cov, data = df.ps, tx = "aodserv") {
  tx <- as.name(tx)
  cov <- as.name(cov)
  data <- data %>%
    group_by(!!tx) %>%
    summarize(
      m = mean(!!cov, na.rm = T),
      s = sd(!!cov, na.rm = T),
      .groups = "drop"
    ) %>%
    pivot_wider(names_from = !!tx, values_from = c("m", "s"), names_sep = "") %>%
    mutate(nd = (m1 - m0) / sqrt(s1^2 + s0^2))
  return(data$nd)
}
```

### Check Balance Using Normalized Difference

Using the normalized difference (Imbens & Wooldridge, 2009), the following covariates have a difference that exceeds .25, indicating that selection bias exists: case status open (`open`), child age group of 0 to 2 years old (`chdage1`), caregiver mental health problem (`mental`), caregiver arrested (`arrest`), caregiver received alcohol or drug (AOD) treatment (`PSH17A`), presence of risk (`ra`), presence of the Composite International Diagnostic Interview–Short Form (CIDI-SF) (`cidi`), and
caregiver report of need (`cgneed`). 

```{r}
# Normalized Differences with .25 Threshold
(covs <- str_extract_all(f, "\\w+")[[3]]) # Vector of Covariates
sapply(covs, nd) %>% round(., 3) %>% sort() # Apply Function "na" on Each Covariate
sapply(covs, nd) %>% .[abs(.) > .25] %>% round(., 3) %>% sort()
```

### Check Balance with Standardized Mean Differences

Several R packages (e.g., `MatchIt` and `cobalt`) that contain functions for balance checking use a similar statistic, the standardized mean difference (SMD). Unlike the normalized difference, the denominator in the SMD is by default the square root of the average of the variances of a variable in the treated and control groups. This use of the pooled standard deviation in the SMD formula is most appropriate when estimating the ATE. When the ATT is desired, the standard deviation of the treated group is most appropriate. When using a more stringent criteria of 0.1 for the SMD, 17 variables do not have satisfactory balance, as seen below:

```{r}
# Standardized Mean Differences with .1 Threshold
(smds <- cobalt::bal.tab(f, data = df.ps, threshold = c(m = .1), s.d.denom = "pooled", 
                binary = "std", abs = T))
cobalt::love.plot(smds)
```

### Function to Check Balance Using Chi-Square Tests

Chi-square tests can also be used to assess balance. However, hypothesis tests are influenced by sample size, which can change due to adjustments such as sample trimming.

```{r}
check_blc <- function(cov, df = df.ps, tx = "aodserv") {
  ct <- broom::tidy(
    chisq.test(
      table(df[[cov]], y = df[[tx]]),
      correct = F
    )
  ) %>%
    add_column(cov = cov, .before = "statistic") %>%
    mutate(p.value = round(p.value, 2))
  return(ct)
}
```

### Check Balance Using Chi-Square Tests

```{r warning=F}
covs2 <- c(
  "married", "educ", "pov", "employ", "open", "race", "chdage",
  "cgage", "CRA47A", "mental", "arrest", "PSH17A", "maltx", "ra",
  "cidi", "cgneed"
)
map_dfr(covs2, check_blc)
```

## Create Subclasses

The crucial condition for a successful propensity score subclassification is to make the propensity score constant within a stratum. Previous studies have shown that a subclassification of five groups efficiently removes 90% of the bias (Cochran, 1968; Rosenbaum & Rubin, 1984). While using more subclasses creates better homogeneity within subclasses and reduces bias in treatment effect estimates, it also results in fewer observations in each subclass and therefore less precise treatment estimates (Du, 1998).

### Get Quintiles

```{r}
# round to 7 digits to replicate Stata code
(q5 <- c(0, round(quantile(df.ps$ps, c(.2, .4, .6, .8), type = 1), 7), 1))
```

### Create Five Subclasses Using Quintiles

Here we sort the propensity score in an ascending order and use the quintiles calculated above to divide the sample into five subclasses. Note that we will preserve the grouping structure with the `dplyr::group_by()` function, which will allow us to carry out additional group-specific operations (e.g., using the function `dplyr::group_map()`).

```{r}
# Cut 5 subclasses
subs <- df.ps %>%
  arrange(ps) %>%
  group_by(q = cut(
    ps,
    include.lowest = F,
    right = F,
    dig.lab = 7,
    breaks = q5,
    labels = F
  ))

# Summary statistics of ps per subclass
subs %>%
  group_by(q) %>%
  summarise(
    obs = n(),
    mean = mean(ps),
    sd = sd(ps),
    min = min(ps),
    max = max(ps), .groups = "drop"
  )
```

## Check Balance After Subclassification

It is important to check balance on the observed covariates between the two groups before proceeding to the next step. The results are not positive. As is often the case, the two groups did not balance on certain covariates, particularly at the two ends of the five subclasses. Note that all participants in Subclass 1 had low propensity scores and that all participants in Subclass 5 had high scores. As such, some covariates show statistical significant differences between the two treatment groups---that is, they were not balanced between treatment conditions; alarmingly, in the highest and lowest subclasses, some participants fell uniquely into one category of certain covariates.

```{r warning=F}
# Check balance on each group in subs
subs %>%
  group_map(~ map_dfr(covs2, check_blc, df = .x))
```

The results of the balance check are summarized as follows, and the summary reports significant differences only: In
Subclass 1, all participants on "risk assessment" fall into the category of "absence," and on "caregiver report of need," they fall into the category of "no need":

```{r}
subs %>% filter(q == 1) %>% count(ra)
subs %>% filter(q == 1) %>% count(cgneed)
```

In Subclass 2, all participants on "risk assessment" fall into the category of "absence," and the chi-square test on "caregiver report of need" is statistically significant (p < .000):

```{r}
subs %>% filter(q == 2) %>% count(ra)
```

In Subclass 3, the chi-square test on "AOD treatment receipt" is statistically significant (p < .01); in Subclass 4, the chi-square test on "caregiver report of need" is statistically significant (p < .05); and in Subclass 5, the chi-square tests on the following covariates are statistically significant: "case status" (p < .05), "caregiver mental health problem" (p < .05), "risk assessment" (p < .001), "CIDI-SF" (p < .01), and "caregiver report of need" (p < .05).

## Trimming

### Search for Best Cutoff Alpha for Overlap

As the above balance checks show, the dataset violates the overlap assumption. In other words, there may be parts of the covariate space with limited number of observations for either the treatment or control group. To address the problem of limited overlap, we can follow Crump et al. (2009) and pursue a trimming strategy. This involves searching for a cutoff value and then discarding participants whose estimated propensity scores are below the cutoff value or above one minus the cutoff value. Alternatively, Crump et al., 2009 suggests that a rule of thumb is to use .1 as the cutoff value and to discard all observations with estimated propensity scores outside the range [0.1, 0.9].

```{r}
# Program to search best cutoff value for overlap (Crum et al., 2009)
trim <- PSweight::PStrim(
  data = as.data.frame(df.ps), # fix for latest package update
  ps.formula = f,
  optimal = T
)

# The rule of thumb of .1 approximates the formula for optimal trimming
trim$delta # or use .079 from the Stata results
1 - trim$delta # or use .921 from the Stata results
```

### Trim Observations

Now that we have found the optimal cutoff point for trimming, we can trim/discard participants whose estimated propensity scores are below .079 and above .921 (i.e., 1 - .079 = .921) from the analysis. This trimming reduces the sample size from 2,723 to 745, or a reduction of 72.6% of the original sample. The sample reduction through trimming is necessary to address the limited overlap problem, but the loss of sample participants is huge. It clearly reduces the study’s external validity. This is a limitation of trimming.

```{r}
summary(df.ps$ps)
df.trimmed <- df.ps %>%
  filter(ps >= .079 & ps <= .921)
summary(df.trimmed$ps)
nrow(df.ps)
nrow(df.trimmed)
nrow(df.ps) - nrow(df.trimmed) # trimmed 1,978 obs
```

## Create Five New Subclasses and Check Balance

Results of the balance check, shown below, are good. Only three covariates remain statistically significant (i.e., "child age" in Subclass 1, "poverty" in Subclass 3, and "CIDI-SF" in Subclass 5), and all other covariates are not statistically significant. Note that the significance of "child age" may be interpreted as chance significance, because it was never significant in all previous schemes.

```{r warning=F}
# Cut 5 subclasses (Type 2 and 5 in quantile() are identical to Stata results)
subs_t <- df.trimmed %>%
  arrange(ps) %>%
  group_by(q = cut(
    ps,
    include.lowest = F,
    right = F,
    dig.lab = 7,
    breaks = c(0, round(quantile(df.trimmed$ps,
      c(.2, .4, .6, .8),
      type = 2
    ), 7), 2),
    labels = F
  ))

# Summary statistics of ps per subclass
subs_t %>%
  group_by(q) %>%
  summarise(
    obs = n(),
    mean = mean(ps),
    sd = sd(ps),
    min = min(ps),
    max = max(ps), .groups = "drop"
  )

# Check balance
subs_t %>%
  group_map(~ map_dfr(covs2, check_blc, df = .x))

# Filter significant results
subs_t %>%
  group_map(~ map_dfr(covs2, check_blc, df = .x) %>% filter(p.value < .05))
```

## Outcome Analysis by Subclass

```{r}
f2 <- as.formula(paste0(
  "Surv(durm, rrpt) ~ aodserv + married + high + bahigh + ",
  "poverty2 + poverty3 + poverty4 + poverty5 + employ + open + black + ",
  "hispanic + natam + chdage1 + chdage2 + chdage3 + CRA47A"
))
subs_t %>%
  group_map(~ broom::tidy(survival::coxph(f2, .x)))
```

### Aggregate Results

Extracting the coefficients of receipt of substance abuse treatment and the corresponding standard errors from all five subclasses, the study then computes the overall effect by aggregating the five estimated coefficients and computes the associated standard error by aggregating the five estimated standard errors. The purpose of this analysis is to discern whether the treatment effect across all five subclasses is statistically significant. 

The weighted average of the within-stratum estimates, given $K$ strata, the sample size per strata $n_k$, the total sample size $N$, and the stratum ATE $\hat\uptau_k$, is:

$$
\hat{\uptau} = \sum^K_{k=1}\frac{n_k}{N}\hat{\uptau_k}
$$

And the variance of the sample ATE is estimated by:

$$
Var(\hat{\uptau}) = \sum^K_{k=1} \left( \frac{n_k}{N} \right)^2 \text{Var} \left[ \bar{Y}_{0k} - \bar{Y}_{1k} \right]
$$

We can then perform a significance test of nondirectional or directional hypothesis via the square root of the variance or the standard error

$$
SE(\hat\uptau) = \sqrt{\text{Var}(\hat\uptau)}
$$

```{r}
# Aggregate the five treatment effects
(subs_coefs <- subs_t %>%
  group_map(~ broom::tidy(survival::coxph(f2, .x)) %>% 
              filter(term == "aodserv") %>% pull(estimate)) %>%
  unlist())
(subs_n <- subs_t %>% count(q) %>% pull(n))
(subs_ate <- sum(subs_coefs * (subs_n / sum(subs_n))))

# Calculate the hazard ratio
exp(subs_ate)

# Aggregate the variance
(subs_vars <- subs_t %>%
  group_map(~ broom::tidy(survival::coxph(f2, .x)) %>% 
              filter(term == "aodserv") %>% pull(std.error)) %>%
  unlist())
(subs_se <- sqrt(sum(subs_vars^2 * (subs_n / sum(subs_n))^2)))

# Calculate z-score and two-sided p-value
(z <- subs_ate / subs_se)
(p.value <- (1 - pnorm(z)) * 2)
```

**Results with Subclassification:** The study finds that by controlling for selection bias, children whose caregivers used substance abuse treatment are more likely to receive a maltreatment re-report, and the hazard of having a re-report for these children is 45.52% higher than that of children whose caregivers did not use substance abuse treatment, although this difference is not statistically significant at the .05 level ($p = .0575$).

**Results without Subclassification:** The Cox regression without controlling for selection bias shows that the hazard ratio of treatment is 1.55 (i.e., exp(0.436) = 1.55), close to that estimated by the subclassified Cox regression, but the finding is statistically significant at the .001 level. The statistical significance of the treatment effect from the "overall" model appears influenced by uncontrolled selectivity:

```{r}
# Cox regression without controlling for selection bias
broom::tidy(survival::coxph(f2, df)) %>% 
  filter(term == "aodserv") %>%
  mutate(hazard.ratio = exp(estimate))
```

## Subclassification with MatchIt

The `MatchIt` package provides a convenient interface for performing subclassification without relying on any other packages.^[https://kosukeimai.github.io/MatchIt/reference/method_subclass.html] Instead of pooling the stratum-specific treatment effects, it produces stratum weights that can be used to estimate a single marginal effect. This approach is known as marginal mean weighting through stratification (MMWS) (Hong, 2010).^[Hong, G. (2010). Marginal Mean Weighting Through Stratification: Adjustment for Selection Bias in Multilevel Data. *Journal of Educational and Behavioral Statistics*, *35*(5), 499–531. https://doi.org/10.3102/1076998609359785] The robust standard error (`robust = T`) must be used because of the MMWS weights.

```{r}
# Without trimming
(mm <- MatchIt::matchit(f, df, method = "subclass", subclass = 5))
table(mm$subclass)
mm$q.cut # vector of propensity score cutpoints to define subclasses
cobalt::love.plot(mm, binary = "std", disp.subclass = T, threshold = c(m = .1))
cobalt::bal.tab(mm, un = T, binary = "std", disp.subclass = T, threshold = c(m = .1))

# With trimmed dataset & outcome analysis with MMWS
(mm2 <- MatchIt::matchit(f, df.trimmed, method = "subclass", subclass = 5))
table(mm2$subclass)
mm2$q.cut # vector of propensity score cutpoints to define subclasses
cobalt::love.plot(mm2, binary = "std", disp.subclass = T, threshold = c(m = .1))
cobalt::bal.tab(mm2, un = T, binary = "std", disp.subclass = T, threshold = c(m = .1))
broom::tidy(survival::coxph(f2, df.trimmed, mm2$weights, robust = T))
exp(0.281)
```

**Results with Subclassification (MatchIt):** The study finds that by controlling for selection bias, children whose caregivers used substance abuse treatment are more likely to receive a maltreatment re-report, and the hazard of having a re-report for these children is 32.45% higher than that of children whose caregivers did not use substance abuse treatment, although this difference is not statistically significant at the .05 level ($p = 0.173$).

\newpage
# Section 5.7: Computer Lab: Running Propensity Score Dosage Analysis with R

## Modeling Doses of Treatment With Multiple Balancing Scores Estimated by a Multinomial Logit Model

In this example, the treatment dosage is the categorized percentage of time a child used Aid to Families with Dependent Children (AFDC) from birth to current age in 1997. The outcome variable is academic achievement. We hypothesize that an increase in levels of using AFDC decreases academic achievement in a linear fashion. 
 
After examining the distribution of percentages of time using AFDC from birth to current age, we create three dose groups: never used, 1% to 33.9% of the time using AFDC from birth to current age in 1997, and 34% to 100% of the time using AFDC from birth to current age in 1997. 

### Load Packages

We need the `nnet` package to run a multinomial logit model.

```{r message=F, warning=F, error=F}
library(nnet)
library(lmtest)
library(sandwich)
```

### Load Data

```{r}
# Load Data
d <- haven::read_dta("data/adv/chpt10.dta")

# Function to replicate Stata tab
tab <- function(df, var) {
  df %>%
    group_by({{ var }}) %>%
    summarise(n = n(), .groups = "drop") %>%
    mutate(
      Percent = 100 * round((n / sum(n)), 4),
      Cum. = 100 * round(cumsum(n / sum(n)), 4)
    ) %>%
    rename(Freq. = n)
}
```

### Create Treatment Variable

In this example, we treated the highest dosage category as the reference group:

```{r message=FALSE, warning=FALSE}
tab(d, k_adc)

# Treatment
d <- d %>%
  mutate(kadc_ds = factor(
    case_when(
      k_adc == 0 ~ 0,
      k_adc > 0 & k_adc < 34 ~ 1,
      k_adc >= 34 ~ 2
    ), levels = c(2, 0, 1)))

tab(d, kadc_ds)
levels(d$kadc_ds)
```

### Generalized Propensity Scores

#### Mulitinomial Logit Model

To investigate the impact of doses on academic achievement, we employ Imbens's (2000) method. A multinomial logit model was first estimated. Variables showing statistically significant differences from bivariate analyses were entered into the multinomial model as predictors.

```{r message=FALSE, warning=FALSE}
# Use multinomial logit model to predict generalized propensity scores
m <- nnet::multinom(
  kadc_ds ~ pcg_adc + age97 + mratio96 + pcged97 + black,
  data = d
)
summary(m)

# Save model predicted probabilities
head(fitted(m))
pr <- fitted(m) %>%
  as_tibble() %>%
  setNames(c("pr2", "pr0", "pr1"))
head(pr)
summary(pr)
```

### Sampling Weights

Note that even though each participant has three GPSs, the GPS for the dose category each participant actually used is chosen for the weighted analysis. Below is the key syntax to create the weight variable. It is important to note that the GPS itself is not the weight; the inverse of a specific GPS is defined as a weight. 

```{r message=FALSE, warning=FALSE}
# define inverse of the generalized p-scores as sampling weights
d <- d %>%
  mutate(dose_w = case_when(
    kadc_ds == 0 ~ 1 / pr$pr0,
    kadc_ds == 1 ~ 1 / pr$pr1,
    kadc_ds == 2 ~ 1 / pr$pr2
  ))
summary(d$dose_w)
```

### Dummy Variables

Next, we create dummy variables for the outcome analysis. For a study with three dosage categories, we create $d - 1 = 3 - 1 = 2$ dummy variables. These variables then are entered into the outcome analysis. In this example, the outcome analysis is a weighted regression.

```{r message=FALSE, warning=FALSE}
# create t-1 dummy variables to measure dose categories
d <- d %>%
  mutate(
    kadc_ds0 = ifelse(kadc_ds == 0, 1, 0),
    kadc_ds1 = ifelse(kadc_ds == 1, 1, 0)
  )

tab(d, kadc_ds0)
tab(d, kadc_ds1)
```

### Regression

#### Without Control

```{r}
# Regression does not control for selection
reg1 <- lm(lwss97 ~ male + black + age97 + pcged97 + mratio96 + pcg_adc + kadc_ds0 +
  kadc_ds1, data = d)
lmtest::coeftest(reg1, vcovCL, cluster = d$pcg_id)
```

#### With Control

The results of the regression with propensity score adjustments show that, other things being equal, (a) "using AFDC 1% to 33.9% of the time" decreases the academic achievement from "never used" by 4.19 units (i.e., $7.48 - 3.29 = 4.19$), (b) "using AFDC 34% to 100% of the time" decreases the academic achievement from "using AFDC 1% to 33.9% of the time" by 3.29 units, and (c) "using AFDC 34% to 100% of the time” decreases the academic achievement from "never used" by 7.48 units. All differences are statistically significant at a .05 level from a one-tailed test. Thus, the results confirm a linear impact of doses of using AFDC on academic achievement: The longer the time a study child was exposed to poverty, the lower his or her academic achievement. Such a conclusion cannot be drawn by analyses treating the child’s use of AFDC as a binary variable. 

```{r}
# Regression controls for selection using inverse p-scores as weights
reg2 <- lm(lwss97 ~ male + black + age97 + pcged97 + mratio96 + pcg_adc + kadc_ds0 +
  kadc_ds1, weights = dose_w, data = d)
lmtest::coeftest(reg2, vcovCL, cluster = d$pcg_id)
```

\newpage
## The Hirano and Imbens Estimator

The package `causaldrf` implements the Hirano and Imbens estimator in R, but some of the arguments are different from those in Stata and therefore yield different results.

### Load Libraries

The package `causaldrf` can be installed using the commented commands:

```{r}
# require(devtools)
# install_version("causaldrf", version = "0.3", repos = "http://cran.us.r-project.org")
library(causaldrf)
```

### Manage Data

For this example, we will be using the same data and research question from above, but instead treat the treatment dosage as a continuous variable, the percentage of time a child used AFDC from birth to 1997.

We first excluded all participants with a zero dosage (i.e., `k_adc == 0`). We then divided the sample into three groups based on the distribution of treatment dosage: [1, 25], [25, 60], and [60, 100]. The three groups are those who used the program from birth to 1997 for 1% to 25% of the time, 25% to 60% of the time, and 60% to 100% of the time. The sample size for each group is 103, 101, and 70, respectively.

```{r data, message=FALSE, warning=FALSE, error=FALSE}
# Load Data
df <- haven::read_dta("data/adv/chpt10.dta")
#stem(df$k_adc)

# Manipulate Data
df2 <- df %>%
  # Exclude participants with zero dosage
  filter(k_adc != 0) %>%
  # Divide sample into three groups based on distribution of treatment dosage
  mutate(cut = ifelse(k_adc <= 25, 25,
               ifelse(k_adc > 25 & k_adc <= 60, 60,
               ifelse(k_adc > 60, 100, NA)))) %>%
  # Rename variables
  rename(
    treat = k_adc,
    Y = lwss97
  )

# Sample Size for Each Group
df2 %>%
  count(cut)

# Calculate the median treatment of each group
grid_val_values <- df2 %>%
  group_by(cut) %>%
  summarize(treat_median = median(treat)) %>%
  pull(treat_median)
grid_val_values
```

### Estimate Median Treatment of Each Group

```{r}
hi_list0 <- causaldrf::hi_est(
  Y = Y,
  treat = treat,
  treat_formula = treat ~ pcg_adc + age97 + male + black + mratio96 + pcged97,
  outcome_formula = Y ~ treat + I(treat^2) + gps + I(gps^2) + treat * gps,
  data = df2,
  grid_val = grid_val_values,
  treat_mod = "LogNormal"
)
hi_sum0 <- summary(hi_list0)
```

### Estimate the Average Dose Response Function (ADRF)

We now moved to the last stage to estimate the dose response function and treatment effect function (i.e., the estimated derivative). Note: we specified a 10-unit increment of the treatment dosage for the x-axis in the plots.

```{r message=FALSE, warning=FALSE, error=FALSE}
hi_list <- causaldrf::hi_est(
  Y = Y,
  treat = treat,
  treat_formula = treat ~ pcg_adc + age97 + male + black + mratio96 + pcged97,
  outcome_formula = Y ~ treat + I(treat^2) + gps + I(gps^2) + treat * gps,
  data = df2,
  grid_val = seq(10, 100, by = 10), # grid_val_values
  treat_mod = "LogNormal"
)
hi_sum <- summary(hi_list)
```

### Plot ADRF

```{r plot, message=FALSE, warning=FALSE, error=FALSE}
adrf <- tibble(
  x = seq(10, 100, by = 10),
  y = hi_list$param
)
ggplot(adrf, aes(x = x, y = y)) +
  geom_line() +
  theme_minimal() +
  ylim(85, 105) +
  scale_x_continuous(breaks = seq(0, 100, by = 20)) +
  labs(
    title = "Average Dose Response Function (ADRF)",
    x = "Treatment Level",
    y = "Outcome"
  )
```

### Balance Check

Below we run unadjusted t-tests for each covariate:

```{r message=F, error=F, warning=F}
# Test Unadjusted mean differences
df3 <- df2 %>%
  mutate(g1 = ifelse(treat <= 25, 1, 0),
         g2 = ifelse(treat > 25 & treat <= 60, 1, 0),
         g3 = ifelse(treat > 60, 1, 0))
df3 %>% count(cut, g1)
df3 %>% count(cut, g2)
df3 %>% count(cut, g3)

unadjust <- function(cov, group) {
  cov_v <- df3 %>% select(all_of(cov)) %>% pull()
  group_v <- df3 %>% select(all_of(group)) %>% pull()
  t.test(cov_v ~ group_v, var.equal = T) %>%
    broom::tidy() %>%
    mutate(Formula = paste0(cov, " ~ ", group)) %>%
    select(Formula, MeanGroup0 = estimate1, MeanGroup1 = estimate2,
           t = statistic, p.value, df = parameter)
}
arg1 <- rep(c("pcg_adc", "age97", "male", "black", "mratio96", "pcged97"), each = 3)
arg2 <- rep(c("g1", "g2", "g3"), 6)
map2_dfr(arg1, arg2, unadjust)
```

Post-adjustment balance checks can be conducted by following the procedure in Galagate & Schafer (2015).^[https://rdrr.io/cran/causaldrf/f/inst/doc/Using_causaldrf.pdf]

\newpage
# Section 6.4: Computer Lab: Running Sensitivity Analysis with rbounds

## Sensitivity Analysis of the Effects

Steps in conducting the sensitivity analysis for matched pair study:

1. Arrange data (i.e., the outcome variable) by pair, list the value of outcome 
variable for the treated and control units separately, and take the difference 
between the two values at the pair level; 

2. Using the Wilcoxon’s signed-rank test to calculate the lower and upper bounds of 
p values when $\Gamma = 1$; 

3. Repeat Step 2 for different values of $\Gamma$; that is, calculate the lower and upper bounds of p values when $\Gamma = 2, 3, \cdots$ until a $\Gamma$ value at which the p value is > .05; this is the $\Gamma$ value at which the significance interval becomes uninformative. The $\Gamma$ value so obtained is the measure of sensitivity to hidden bias. A large $\Gamma$ is good, meaning that the study is insensitive to hidden selection bias. 

#### Load Packages

We will use the package `rbounds` for the sensitivity analysis and `coin` for the Wilcoxon signed-rank test.

```{r message=F, warning=F, error=F}
library(rbounds)
library(coin)
library(kableExtra)
```

#### Load Data

The data below contains blood lead levels (ug/dl) of children whose parents are exposed to lead at their places of work (`trt`), versus children whose parents are unexposed to lead at their places of work (`ctrl`):

```{r message=FALSE, warning=FALSE}
# Data: Matched Data of Lead Blood Levels in Children
# Source: Rosenbaum (2002a), p. 82
df.sa <- tibble(
  trt = c(
    38, 23, 41, 18, 37, 36, 23, 62, 31, 34, 24, 14, 21, 17, 16, 20, 15, 10, 45,
    39, 22, 35, 49, 48, 44, 35, 43, 39, 34, 13, 73, 25, 27
  ),
  ctrl = c(
    16, 18, 18, 24, 19, 11, 10, 15, 16, 18, 18, 13, 19, 10, 16, 16, 24, 13, 9,
    14, 21, 19, 7, 18, 19, 12, 11, 22, 25, 16, 13, 11, 13
  )
) %>%
  rowid_to_column(var = "pair") %>%
  mutate(diff = trt - ctrl)
head(df.sa)
```

Below, we see the range of significance levels for the test statistic when $\Gamma$ is equal to various values. As depicted, $\Gamma = 4.25$ is the value at which the significance interval becomes uninformative. Rosenbaum (2002) interpreted this finding in the following way: "The table shows that to explain away the observed association between parental exposure to lead and child’s lead level, a hidden bias or unobserved covariate would need to increase the odds of exposure by more than a factor of $\Gamma = 4.25$" (p. 115).

```{r message=FALSE, warning=FALSE}
p <- rbounds::psens(df.sa$trt, df.sa$ctrl, GammaInc = 0.01)
p$bounds %>%
  filter(Gamma %in% c(1:4, 4.25, 5:6))
```

## Section 11.5.2 Sensitivity Analysis & Pair Matching

A study using regression analysis of difference scores found that in 1997, children who used AFDC had an average letter-word identification score that was 3.17 points lower than that of children who never used AFDC ($p < .05$). Furthermore, the study adjusted selection based on six observed variables (i.e., the ratio of family income to poverty threshold in 1996, caregiver’s education in 1997, caregiver’s history of using welfare, child race, child age in 1997, and child gender). We now examine the same data in an attempt to determine to what extent is this study sensitive to hidden selection bias. 
 
Results of the sensitivity analysis are shown below. Using the Wilcoxon’s signed-rank test, the sensitivity analysis shows that the study becomes sensitive to hidden bias at $\Gamma = 1.43$. Because 1.43 is a small value, we can conclude that the study is very sensitive to hidden bias, and therefore, further analysis that controls for additional biases is warranted.

```{r message=FALSE, warning=FALSE}
# Import data and pivot data from long to wide
d <- haven::read_dta("data/adv/chpt5_2.dta") %>%
  select(pm, kuse, lwss97) %>%
  drop_na() %>%
  filter(pm != "NA") %>%
  arrange(pm, kuse) %>%
  group_by(pm) %>%
  mutate(kuse = recode(kuse, `0` = "y0", `1` = "y1")) %>%
  pivot_wider(names_from = kuse, values_from = lwss97) %>%
  ungroup()
```

```{r message=FALSE, warning=FALSE}
p1 <- rbounds::psens(d$y0, d$y1, GammaInc = 0.01)
p1$bounds %>%
  filter(Gamma %in% c(1, 1.3, 1.42, 1.43, 1.5, 2))
```

We use the Wilcoxon signed rank test to test the significance of y. The results show that the randomization test assuming no hidden bias is highly significant ($z = -4.191$, $p < .001$). That is, there is a strong treatment effect. 

```{r message=FALSE, warning=FALSE}
coin::wilcoxsign_test(y1 ~ y0, data = d)
```

We can further compare the level of sensitivity to bias of the AFDC study to that of other studies. The table below presents results of sensitivity analyses for four observational studies (Rosenbaum, 2005, table 4). Among the four studies, Study 2 is the least sensitive to hidden bias; the study of the effects of diethylstilbestrol becomes sensitive at about $\Gamma = 7$. In contrast, Study 4 is the most sensitive to hidden bias; the study of the effects of coffee becomes sensitive at about $\Gamma = 1.3$. 

```{r echo=F}
tibble(
  Treatment = c(
    "Study 1: Smoking/Lung Cancer",
    "Study 2: Diethylstilbestrol/Vaginal Cancer",
    "Study 3: Lead/Blood Lead",
    "Study 4: Coffee/Myocardial Infarction"
  ),
  `Gamma = 1` = c(
    "< .001", "< .001", "< .001", ".0038"
  ),
  `Gamma, max p-value` = c(
    "(5, 0.03)", "(7, 0.054)", "(4.25, 0.047)", "(1.3, 0.056)"
  )
) %>%
  kbl() %>%
  kable_paper()
```

Rosenbaum (2005) provided the following explanation of the implications of the sensitivity analyses for these studies: "A small bias could explain away the effects of coffee, but only an enormous bias could explain away the effects of diethylstilbestrol. The lead exposure study, although quite insensitive to hidden bias, is about halfway between these two other studies, and is slightly more sensitive to hidden bias than the study of the effects of smoking" (p. 1812).

Compared with these four studies, our study of the effect of welfare use on academic achievement is slightly better (i.e., less sensitive to unobserved bias) than the study of the effects of coffee, but it is more sensitive to hidden bias than the three other studies.